Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-10-08 manuscript/manuscript.Rmd
Abstract

---
title: "Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination"
shorttitle        : "Flattering Advice"
author: 
  - name          : "Amanda Chen"
    affiliation   : "1"
    corresponding : yes    
    email         : "zchengj@connect.ust.hk"
    address       : "Department of Management, The Hong Kong University of Science and Technology, Hong Kong"
  - name          : "David Hagmann"
    affiliation   : "1"
affiliation:
  - id            : "1"
    institution   : "The Hong Kong University of Science and Technology"
note : |
  `r format(Sys.time(), "%e %B, %Y")`
abstract: |
  The purpose of advice is typically understood as helping recipients make better decisions and solve problems. However, in this paper, we propose that advisors also consider the recipient's belief utility and may try to avoid disappointing them, leading to what we term "flattering advice." Across three pre-registered experiments involving real interactions between advisors and advisees (n = 3,900), we demonstrate that individuals factor in others' expectations when giving advice, even when those expectations are not relevant to solving the decision-making problem. Specifically, advisors take advisees' expectations into account, tailoring their advice to create a more positive impression of the advisee and avoid causing disappointment. (Study 1). Advisors inflate their advice due to interpersonal concerns, with those who care about being liked by the advisee inflating their advice to a greater extent. Advisees, in turn, reward this flattery, perceiving these advisors as more likable and equally trustworthy compared to those who give more accurate advice (Study 2). Additionally, when advisors match advisees' expectations, men tend to receive more aspirational advice than women, due to their higher confidence, even when only expectations (not gender) are visible to the advisor. However, relying on these expectations leads to poorer decision outcomes (Study 3). We conclude by discussing the practical and theoretical implications of our findings, as well as directions for future research.

>>>>

---
title: "Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination"
shorttitle        : "Flattering Advice"
author: 
  - name          : "Amanda Chen"
    affiliation   : "1"
    corresponding : yes    
    email         : "zchengj@connect.ust.hk"
    address       : "Department of Management, The Hong Kong University of Science and Technology, Hong Kong"
  - name          : "David Hagmann"
    affiliation   : "1"
affiliation:
  - id            : "1"
    institution   : "The Hong Kong University of Science and Technology"
note : |
  `r format(Sys.time(), "%e %B, %Y")`
abstract: |
  The primary purpose of advice is often seen as helping recipients make better decisions and solve problems. However, this paper reveals that advisors are also influenced by interpersonal concerns. Specifically, we propose that advisors consider the recipient's "belief utility" and may seek to avoid disappointing them, resulting in what we call "flattering advice." Through three pre-registered experiments involving real interactions between advisors and advisees (n = 3,900), we demonstrate that advisors account for advisees' expectations, even when irrelevant to solving the problem. Specifically, advisors tailor their advice to match advisees' expectations about their own capabilities, aiming to avoid disappointing them (Study 1). Advisors inflate their advice due to interpersonal concerns, with those who care about being liked by the advisee inflating their advice to a greater extent. Advisees, in turn, reward this flattery, perceiving these advisors as more likable and equally trustworthy compared to those who aim to give more accurate advice (Study 2). Additionally, when advisors match advisees' expectations, men tend to receive more aspirational advice than women, due to their higher confidence, even when only expectations (not gender) are visible to the advisor. However, relying on these expectations leads to poorer decision outcomes (Study 3). We conclude by discussing the practical and theoretical implications of our findings, as well as directions for future research.
